---
title: "Intro to R Workshop"
format: html
editor: visual
author: "Katie Bunge"
date: 2026-03-18
---

I am learning R. This is my markdown segment.

Setup Chunk with Package Installation is included below:

```{r}
#| label: setup
source("training_source.R")

```

Examples of Numeric Values

```{r}
#| label: numeric values
#This is a comment within the code segment. 

42
194
typeof(239.2)

```

Examples of Character Values

```{r}
#These are examples of character values. 

"This is a character string"
"SLIPJIG"
typeof("2/4 is a polka")

```

Working with Vectors

```{r}

#The operator 'c' stands for 'concatenate'

numbervector <- c(1,2,3)
charactervector <- c("a", "b", "c")
binaryvector <- c(TRUE, FALSE)
mixedvector <- c(1, "two", 3)

numbervector
charactervector
binaryvector
mixedvector

```

Exploring Lists

```{r}
my_list <- list(numbers = c(1,2,3), words = c("a","b"), mixed = list(TRUE, 5))
str(my_list)



```

Data Frames

```{r}
sampledf <- data.frame(id = 1:3,
                       mode = c("walk","car","bus"),
                       distance = c(1.2, 5.0, 3.4))
sampledf 

```

Importing and Viewing Data

Import = load data from a file Cleaning = prepare column names Viewing = inspect rows and column types

Importing data with rio

```{r}
csv_path <- "../data/raw/presurvey_20260317.csv"
survey_raw <- rio::import(csv_path)
print(survey_raw)
```

-- BEGIN DAY 2 OF TRAINING --

```{r}
survey_raw |> 
  glimpse()

survey <- survey_raw |> 
  clean_names() |> 
  rename(os = what_operating_system_are_you_using_for_this_workshop,
         prior_exp = which_of_the_following_are_true_regarding_your_experiences_with_r_prior_to_this_workshop,
         r_hope = what_do_you_hope_to_use_r_for_in_the_future,                                                 travel_mode = the_workshop_will_be_held_on_the_psu_campus_what_is_your_intended_mode_of_travel_to_the_workshop_check_all_that_apply,
         travel_direct = do_you_intend_to_travel_directly_from_home_to_the_workshop,
         distance = how_far_will_you_likely_travel,                               
         commute_time = what_is_the_approximate_travel_time_for_your_morning_commute_to_psu,                   drink = what_is_your_morning_drink_of_choice,                                                         movie_quote = what_is_the_first_movie_quote_that_comes_to_mind,                                       pineapple_ok = pineapple_is_an_acceptable_pizza_topping
         )

survey |> 
  glimpse()

```

```{r}
linux <- c("linux", "ubuntu", "debian", "fedora", "arch")

survey2 <- survey |> 
  mutate(commute_digits = str_extract(commute_time, "[0-9]+"),
         distance_digits = str_extract(distance, "[-]?[\\d]*\\.?[\\d]+"),
         commute_min = as.numeric(commute_digits),
         distance_mi = as.numeric(distance_digits),
         travel_mph = if_else(!is.na(commute_min) & commute_min > 0,
                              distance_mi / commute_min * 60, NA_real_),
         os = str_to_lower(os),
         os_family = case_when(str_detect(os, "win") ~ "Windows",
                               str_detect(os, "mac|apple") ~ "Mac",
                               str_detect(os, paste(linux, collapse = "|")) ~ "Linux",
                               TRUE ~ "Other"
                               )
         )

print(survey2)

survey3 <- survey2 |> 
  select(os, os_family, distance_mi, commute_min, travel_mph)

head(survey3)

```

Data in a long versus wide format 

Example: Survey Data 

Wide Format 

* Q1, Q2, Q3, Q4 are separate variables or "columns" -> this is a "wide" format 
Row 1: D1.1, D1.2, D1.3, D1.4 
Row 2: D2.1, D2.2, D2.3, D2.4 

Long Format

Data | Variable 
--------------
D1.1 |  Q1 
D1.2 |  Q2 
D2.2 |  Q2 

@>->->-| Working to Transform Data   |-<-<-@  


```{r}
long_survey <- survey2 |> 
  select(distance, commute_time) |> 
  pivot_longer(cols = c(distance, commute_time), 
               names_to = "num_text_var", 
               values_to = "numeric_text") |> 
  mutate(numeric_text = str_to_lower(numeric_text), .rid = row_number())
  
  wide_survey <- long_survey |> 
    pivot_wider(id_cols = .rid, 
                names_from = num_text_var, 
                values_from = numeric_text, 
                names_repair = "minimal") |> 
    select(-.rid)

print(wide_survey)
print(long_survey)
```

```{r}


```


```{r}


```

